---
title: "Class 12"
subtitle: "Bioinformatics in drug discovery and design"
author: "Yi Fu"
date: "5/9/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, let's check if "bio3d" package is installed. And then, load the package.
```{r include=FALSE}
if (!require("bio3d")) {
  install.packages("bio3d")
}
```

```{r}
library(bio3d)
```

And download *Autodock Vina* [here](http://vina.scripps.edu/download.html). 

## 1. Autodock Vina
For this section, we will use the program called Autodock Vina [4]. Autodock Vina is a fast docking program that requires minimal user intervention and is often employed for highthroughput virtual screening. We will run it from the command line.

Now we are ready to run autodoc vina to perform the docking. We will keep a log of all program
output in a file log.txt
```{bash}
#~/Downloads/autodock_vina_1_1_2_mac/bin/vina --config config.txt --log log.txt
```

Once the run is complete, you should have two new files *all.pdbqt*, which contains all the
docked modes, and *log.txt*

## 2. Process the docking results with Bio3D in R
```{r}
res <- read.pdb("all.pdbqt", multi=T)
write.pdb(res, "results.pdb")
```

Let's calculate the RMSD (root mean square distance) between each of the docking results and the known crystal structure using the bio3d package.
```{r}
ori <- read.pdb("1hsg_ligand.pdbqt")
rmsd(ori, res)
```

Normal mode analysis (NMA) is one of the major simulation techniques used to probe largescale motions in biomolecules. Typical application is for the prediction of functional motions in
proteins.
```{r}
pdb <- read.pdb("1HEL")
modes <- nma(pdb)
plot(modes, sse=pdb)
mktrj(modes, mode=7, file="nma_7.pdb")
```

