---
title: "Class 14"
subtitle: "The analysis of RNA-Seq data"
author: "Yi Fu"
date: "5/16/2019"
output: html_document
#always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, let's check if "DESeq2" package is installed. And then, load the package.
```{r include=FALSE}
if (!require("AnnotationDbi")) {
  BiocManager::install("AnnotationDbi")
}
if (!require("org.Hs.eg.db")) {
  BiocManager::install("org.Hs.eg.db")
}
if (!require("DESeq2")) {
  BiocManager::install("DESeq2")
}
if (!require("ggplot2")) {
  install.packages("ggplot2")
}
```

```{r}
library(AnnotationDbi)
library(org.Hs.eg.db)
library(DESeq2)
library(ggplot2)
```

## 1. Processing the data
The computational analysis of an RNA-seq experiment begins from the FASTQ files that contain the nucleotide sequence of each read and a quality score at each position. These reads must first be aligned to a reference genome or transcriptome. The output of this alignment step is commonly stored in a file format called SAM/BAM. This is the workflow we followed last day.

Lets perform some exploratory differential gene expression analysis. **Note: this analysis is for demonstration only. NEVER do differential expression analysis this way!**

First, load the data.
```{r}
counts = read.csv("data/airway_scaledcounts.csv", stringsAsFactors = FALSE,row.names = 1)
metadata = read.csv("data/airway_metadata.csv", stringsAsFactors = FALSE)
rmarkdown::paged_table(counts)
rmarkdown::paged_table(metadata)
```

Subset the **metadata** into two data frames.
```{r}
control = metadata[metadata$dex=="control",]
treated = metadata[metadata$dex=="treated",]
rmarkdown::paged_table(control)
rmarkdown::paged_table(treated)
```

Determine the mean count values for all genes accross control experiments
```{r}
control.mean = rowSums(counts[,control$id])/length(counts[,control$id])
treated.mean = rowSums(counts[,treated$id])/length(counts[,treated$id])
```

Let's store the control.mean and treated.mean together for ease of use
```{r}
meancounts = data.frame(control.mean, treated.mean)
rmarkdown::paged_table(meancounts)
colSums(meancounts)
```

Plot the control vs treated.
```{r}
plot(meancounts[,1:2],xlab="Control",ylab="Treated")
```

We see a few dozen dots outside of the big clump around the origin. So, let's try plotting both axes on a log scale.
```{r}
plot(meancounts[,1:2],log="xy",xlab="log Control",ylab="log Treated")
```

Here we calculate log2foldchange.
```{r}
meancounts$log2fc = log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
```

We will now remove the NaN and -Inf values.
The NaN is returned when you divide by zero and try to take the log. The -Inf is returned when you try to take the log of zero.
```{r}
zero.vals = which(meancounts[,1:2]==0, arr.ind=TRUE)
to.rm = unique(zero.vals[,1])
meancounts = meancounts[-to.rm,]
rmarkdown::paged_table(meancounts)
```

How many genes are up in the drug treated cells and how many are down?
```{r}
up.ind = meancounts$log2fc > 2
down.ind = meancounts$log2fc < (-2)
sum(up.ind)
sum(down.ind)
```

## 2. Annotation

We can add annotation from a supplied CSV file, such as those available from ENSEMBLE or UCSC. The annotables_grch38.csv annotation table links the unambiguous Ensembl gene ID to other useful annotation like the gene symbol, full gene name, location, Entrez gene ID, etc.
```{r}
anno = read.csv("data/annotables_grch38.csv")
```

Use the **merge** funtion to add the annotation data from the "anno" object to our RNA-Seq results in "meancounts"
```{r}
# use the merge function
meancounts.anno = merge(meancounts, anno, by.x="row.names", by.y="ensgene")
rmarkdown::paged_table(meancounts.anno)
```

We can use the *mapIds()* function to add individual columns to our results table.
```{r}
columns(org.Hs.eg.db)
meancounts$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(meancounts),
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")
meancounts$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(meancounts),
                     column="ENTREZID",
                     keytype="ENSEMBL",
                     multiVals="first")
meancounts$uniprot <- mapIds(org.Hs.eg.db,
                     keys=row.names(meancounts),
                     column="UNIPROT",
                     keytype="ENSEMBL",
                     multiVals="first")
```

```{r}
rmarkdown::paged_table(meancounts[up.ind,])
rmarkdown::paged_table(meancounts[down.ind,])
```

## 3. Differential gene expression

DESeq is Differential gene expression analysis based on the negative binomial distribution. Setup the object needed for DESeq analysis
```{r}
counts <- read.csv("data/airway_scaledcounts.csv", stringsAsFactors = FALSE)
dds = DESeqDataSetFromMatrix(countData=counts, 
                             colData=metadata, 
                             design=~dex, 
                             tidy=TRUE)
dds = DESeq(dds)
```

```{r}
res = results(dds)
summary(res)
res = res[order(res$pvalue),]
rmarkdown::paged_table(as.data.frame(res))
write.csv(res, file="data/deseq_results.csv")
```

Change the p-value threshold again to 0.01
```{r}
res01 = results(dds, alpha=0.01)
summary(res01)
```

Store our results as a data.frame object, sort or order our results by the adjusted p-value
```{r}
resSig01 = subset(as.data.frame(res), padj < 0.01)
rmarkdown::paged_table(resSig01[order(resSig01$padj),])
```


## 4. Data visualization

Let’s first see what the gene ID is for the CRISPLD2 gene using:
```{r}
resSig01[grep("CRISPLD2", resSig01$symbol),]
rownames(resSig01[grep("CRISPLD2", resSig01$symbol),])
```

We can mow use this returned object to plot a boxplot with the base graphics function
```{r}
d = plotCounts(dds, gene="ENSG00000103196", intgroup="dex", returnData=TRUE)
head(d)
boxplot(count ~ dex , data=d)
```

Or use **ggplot2**
```{r}
ggplot(d, aes(dex, count)) + 
geom_boxplot(aes(fill=dex)) + 
scale_y_log10() + 
ggtitle("CRISPLD2")
```

Volcano plot
```{r}
# Setup our custom point color vector 
mycols = rep("gray", nrow(res))
mycols[ res$padj < 0.01 ]  <- "red" 
mycols[ abs(res$log2FoldChange) > 2 ]  <- "blue"
mycols[ (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 ) ] <- "green"

palette( c("gray","blue") )
plot( res$log2FoldChange,  -log(res$padj), col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.1), col="darkgray", lty=2)
```

Or use **ggplot2**
```{r}
ggplot(as.data.frame(res), aes(log2FoldChange, -log10(pvalue), col=mycols)) + 
geom_point() + 
ggtitle("Volcano plot")
```

