---
title: "Class 4: Bioinformatics data analysis with R"
author: "Yi Fu"
date: "4/11/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Simple Calculations
Example of Addition:
```{r}
5 + 3
```

Example of Subtraction:
```{r}
5 - 3
```

Example of Multiplication:
```{r}
5 * 3
```

Example of Division:
```{r}
5 / 3
```


## 2. Object Assignment

Examples of object Assignment:
```{r}
x = 3 * 4
x <- 3 * 4
```

## 3. Function Calling
Examples of calling *seq()* function:
```{r}
seq(1, 10)
seq(1, 10, by=2)
```

Example of calling *date()* function:
```{r}
date()
```

## 4. Getting help
When we don't remember what a function does and what are the parameters to take in, we can use *help()* or *?* to check.
```{r}
help(seq)
?seq
```

We can also take a look at demos of the function.
```{r}
example(seq)
```

## 5. R Vectors
Let's play with vectors in R, with the following two vectors *x* and *y*.
```{r}
x = c(56, 95.3, 0.4)
length(x)
y = c(3.2, 1.1, 0.2)
length(y)
```

Here are the vector arithmetics:
```{r}
x + y
x - y
x * y
x / y
sqrt(x)
round(sqrt(x), 3)
log(x) / 2 + 1
```

Get vector values:
```{r}
x[1]
x[4]
```

Set vector values:
```{r}
x[3] = 0.5
x[4] = 1
```

Now, x is a vector of length 4.
```{r}
x
```


## RStudio Shortcuts
Use **Alt+Shift+K** to see shortcut quick reference. More shortcuts can be found [here](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts).


### Some Useful Shortcuts
```{bash eval=FALSE}
Command+Shift+K    # knit document
Command+Option+I   # Insert chunk (Sweave and Knitr)
Command+Enter      # Run current line/selection
Command+Option+R   # Run the current document
Command+Option+B   # Run from document beginning to current line
Command+Shift+C    # comment and uncomment
```
