---
title: "Homework for class 06"
author: "Yi Fu"
date: "Due on 4/25/2019"
output: html_document
---
## Q6. How would you generalize the original code above to work with any set of input protein structures?

First, let's check if "bio3d" package is installed. And then, load the package.
```{r results='hide', warning=F}
## install "bio3d" package
if (!require("bio3d")) {
  install.packages("bio3d")
}
```

### Question: Can you improve this analysis code?
Run through the code to see if it works, fix any copy/paste errors before simplifying to a core working code snippet, reducing any calculation duplication. And then, find a way to improve the code by writing a new function.
```{r eval=F}
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug

s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s3, chain="A", elety="CA")  ### changed the typo "s1" to "s3"

s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b

plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

### Answer: code for the analysis of protein drug interactions by abstracting the main activities in one more useful function call.
```{r}
library("bio3d")

# The function takes a vector of pdbid's and output the plot of factor of corresponding chain and atoms. 
# The pdbid is required parameter. 
# Usage: 
## plotPDB (pdbid, ...)
# Parameters:
## pdbid: a vector of 4-character identifiers of entries in the PDB
## chain: which chain to look at (chain A is default)
## elety: which atom names to look at (C-alpha is default)
plotPDB <- function (pdbid, chain="A", elety="CA") {

  # examine each id one by one
  for (protein in pdbid) {
    
    # access on-line PDB file and get the PDB structure object
    s <- read.pdb(protein)
    
    # get a subset of the PDB structure object, given chain and elety
    s.chain <- trim.pdb(s, chain=chain, elety=elety)
    
    # get bfactor: the displacement of the atomic positions from an average (mean) value.
    s.b <- s.chain$atom$b
    
    # outputs a plot: Bfactor vs Residue
    plotb3(s.b, sse=s.chain, typ="l", ylab="Bfactor", main=protein)
  }
}
```

Let's call the function to check the function we just created is working. This function calling plots for three protein including "4AKE", "1AKE", "1E4Y".
```{r}
plotPDB (c ("4AKE", "1AKE", "1E4Y"))
```