---
title: "Class 7"
subtitle: "More on R functions"
author: "Yi Fu"
date: "4/23/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Note: the [original](https://tinyurl.com/rescale-R) *both_na*, *gene_intersect* functions

## 1. *both_na* funtion

**Start with a simple example of the larger problem I am trying to solve**
```{r}
x = c(1 ,2, NA, 3, NA)
y = c(NA, 3, NA, 3, 4)
```

```{r}
is.na(x)
is.na(y)
is.na(x) & is.na(y)
sum(is.na(x) & is.na(y))
which (c (F, F, T, F, T))
which (is.na (c (1, 2, NA, 4)))
```

**Now, I can make this into a function.**

### Version 1:
Check how many NA entries in both input vectors.
```{r}
both_na1 <- function (x, y) {
  sum (is.na (x) & is.na (y))
}
```

Examples:
```{r}
both_na1 (c(1 ,2, NA, 3, NA), c(NA, 3, NA, 2, NA))
both_na1 (c(NA, NA, NA), c(1, NA, NA))
both_na1 (c(NA, NA, NA), c(1, NA, NA, NA, 1, NA)) # both_na(c(NA, NA, NA,NA, NA, NA), c(1, NA, NA, NA, 1, NA))
```

### Version 2:
Version 1 + check the length of two vectors.
```{r}
both_na2 <- function(x, y) {
  if(length(x) != length(y)) {
    print("Input x and y should be vectors of the same length")
  }
  sum(is.na(x) & is.na(y))
}
```

Example:
```{r}
both_na2 (c(NA, NA, NA), c(1, NA, NA))
both_na2 (c(NA, NA, NA), c(1, NA, NA, NA, 1, NA))
```

### Version 3:
Version 2 + check NA position(s).
```{r}
both_na3 <- function(x, y) {
  if(length(x) != length(y)) {
    print("Input x and y should be vectors of the same length")
  }
  na.in.both = is.na(x) & is.na(y)
  na.number = sum(na.in.both)
  na.which = which(na.in.both)

  message("Found ", na.number, " NA's at position(s):", paste(na.which, collapse=", ")) 
  
  return(list(number=na.number, which=na.which))
}
```

Examples:
```{r}
both_na3 (c(NA ,2, NA, 3, NA), c(NA, 3, NA, NA, 4))
```

## 2. *gene_intersect* function

**Start with a simple example of the larger problem I am trying to solve**
```{r}
df1 = data.frame(IDs=c("gene1", "gene2", "gene3"),
                 exp=c(2,1,1),
                 stringsAsFactors=FALSE)
df2 = data.frame(IDs=c("gene2", "gene4", "gene3", "gene5"),
                 exp=c(-2, NA, 1, 2),
                 stringsAsFactors=FALSE)
x = df1$IDs
y = df2$IDs
```

```{r}
intersect(x, y)
x %in% y
x[x %in% y]
y %in% x
y[y %in% x]
cbind( x[ x %in% y ], y[ y %in% x ] )
```

**We can use RStudio shortcut 'CODE > EXTRACT FUNCTION' to turn our snippet into a working function**

### Version 1:

Find the intersection of two gene lists.
```{r}
gene_intersect1 <- function(x, y) {
  cbind( x[ x %in% y ], y[ y %in% x ] )
}
```

Example:
```{r}
gene_intersect1 (df1$IDs, df2$IDs)
```

**It seems to contain duplicated information. Let's improve that.**

### Version 2:

Version 1 + get values for each gene.
```{r}
gene_intersect2 <- function(df1, df2) { 
   cbind( df1[ df1$IDs %in% df2$IDs, ], 
          df2[ df2$IDs %in% df1$IDs, "exp"] )
}
```

Example:
```{r}
gene_intersect2 (df1, df2)
```

### Version 3:

Version 2 + specify the name of gene column.
```{r}
gene_intersect3 <- function(df1, df2, gene.colname="IDs") { 
   cbind( df1[ df1[,gene.colname] %in% df2[,gene.colname], ], 
          exp2=df2[ df2[,gene.colname] %in% df1[,gene.colname], "exp"] )
}
```

Example:
```{r}
gene_intersect3 (df1, df2)
```

### Version 4:

Neat version of Version 3.
```{r}
gene_intersect4 <- function(df1, df2, gene.colname="IDs") { 

  df1.name <- df1[,gene.colname]
  df2.name <- df2[,gene.colname]

  df1.inds <- df1.name %in% df2.name
  df2.inds <- df2.name %in% df1.name

   cbind( df1[ df1.inds, ], 
          exp2=df2[ df2.inds, "exp"] )
}
```

Example:
```{r}
gene_intersect4 (df1, df2)
```

## 3. *score* function

dropped the lowest score
```{r}
score <- function(x) {
  (sum (x) - min (x))/ (length (x) - 1)
}
```

Examples:
```{r}
score (c (100, 100, 100, 100, 100, 100, 100, 90))
score (c (100, 90, 90, 90, 90, 90, 97, 80))
```

